---
title: "Gantt Style Project Overview"
output:
  html_document: default
  html_notebook: default
---

# Introduction

The purpose of this notebook are to firstly to generate realistic project milestone data  for a large project (similar to a BCBS-239 programme) then secondly to build an example Gantt chart 
but one that shows the details as well as an overview.

```{r echo=FALSE}
if ("tidyr" %in% rownames(installed.packages()) == FALSE) {install.packages("tidyr", repos = "http://cran.ma.imperial.ac.uk/")}
if ("dplyr" %in% rownames(installed.packages()) == FALSE) {install.packages("dplyr", repos = "http://cran.ma.imperial.ac.uk/")}
if ("readr" %in% rownames(installed.packages()) == FALSE) {install.packages("readr", repos = "http://cran.ma.imperial.ac.uk/")}
if ("ggplot2" %in% rownames(installed.packages()) == FALSE) {install.packages("ggplot2", repos = "http://cran.ma.imperial.ac.uk/")}
if ("forcats" %in% rownames(installed.packages()) == FALSE) {install.packages("forcats", repos = "http://cran.ma.imperial.ac.uk/")}
if ("scales" %in% rownames(installed.packages()) == FALSE) {install.packages("scales", repos = "http://cran.ma.imperial.ac.uk/")}
if ("stringr" %in% rownames(installed.packages()) == FALSE) {install.packages("stringr", repos = "http://cran.ma.imperial.ac.uk/")}
if ("openxlsx" %in% rownames(installed.packages()) == FALSE) {install.packages("openxlsx", repos = "http://cran.ma.imperial.ac.uk/")}
```

```{r  warning=FALSE, message=FALSE}
library(tidyr)
library(dplyr)
library(readr)
library(ggplot2)
library(forcats)
library(stringr)
library(scales)
library(openxlsx)
```

Create the basic parameters to generate some realistic random data.

```{r message=FALSE}
# This csv contains the 11 principles 
(df_p <-readr::read_csv("Principle.csv"))
```

```{r}

# There are 11 principle with keys 1 through 11.  The next lines generate a vector 
# of principle keys contain 30 Principle 1, 35 Principle 2 and so on
# This vector is used in a sample function to generate milestones.

# Overall, principles 1 and 2 have many milestones, 10 and 11 have relatively fewer
principle_dist <- c(rep(1:11, c(30, 35, 10, 7, 8, 17, 22, 10, 10, 5, 3)))

# This distribution will generate relatively more milestones for principle 6 and 9
principle_dist1 <- c(rep(1:11, c(10, 11, 10, 7, 45, 17, 22, 40, 10, 5, 2)))

# This distribution will generate relatively more milestones for principle 2, 7 and 9
principle_dist2 <- c(rep(1:11, c(5, 35, 10, 7, 8, 2, 22, 10, 30, 5, 1)))

# The project runs from 2016 to 2018 inclusive.  

project_length_days = 365 * 3 # 3 years
project_start_date <- as.Date("2016-01-01")

project_critical_date1 <- as.Date("2016-12-05")
project_critical_date2 <- as.Date("2017-12-01")

current_date <- as.Date("2017-09-12")

# This give a distribution for the relative weights (importance) of the milestones
weight_values <- c(0.1, 0.3, 0.5, 0.9, 2.7, 5)

milestone_status_values <- c("Active", "Completed")

```

```{r}
# Generate a baseline of milestones evenly spread over the project timescale
n <- 1000
df <- data_frame(
  PrincipleKey = sample(principle_dist, n, replace = TRUE),
  MilestoneDate = sample(project_start_date + 1:project_length_days, n, replace = TRUE)
)

# Generate a overlay of milestones clustered around critical_date_1 and with a different distribution over the principles
n <- 500
df_bump1 <- data_frame(
  PrincipleKey = sample(principle_dist1, n, replace = TRUE),
  MilestoneDate = project_critical_date1 + floor(rnorm(n, 0, 50))
)

# Generate another overlay of milestones clustered around critical_date_2 and also with a different distribution over the principles
n <- 600
df_bump2 <- data_frame(
  PrincipleKey = sample(principle_dist2, n, replace = TRUE),
  MilestoneDate = project_critical_date2 + floor(rnorm(n, 0, 30))
)

# Combined to get the final set of milestone data
df_total <- bind_rows(df, df_bump1, df_bump2)

# Add status information (whether active, complted or overdue) to the milestones.
# Firstly, assign an active or completed status based on the current_date 
# (using rnorm to allow some leeway so some future milesones are completed and some past one are active)
# Then update all historic milestones with active status to overdue status.
n_total <- nrow(df_total)
df_total <-
  df_total %>%
  mutate(
    MilestoneKey = 1:n_total,
    MilestoneStatus = ifelse(MilestoneDate > current_date + rnorm(n_total, 0, 120), "Active", "Completed"),
    MilestoneStatus = ifelse(MilestoneDate < current_date & MilestoneStatus == "Active", "Overdue", MilestoneStatus),
    Weight = sample(weight_values, n_total, replace = TRUE)
  )

# Join on the principle name and remove the surrogate key to get the visual-ready dataset.
df_final <- df_total %>%
  inner_join(df_p, by = "PrincipleKey") %>%
  mutate(PrincipleName = fct_reorder(PrincipleName, -PrincipleKey))

df_final
```

Create a visual that shows all milestones by principle over the three year period of the project. Base on a Gantt chart so it is instantly familar to project managers.  Use jittering and partially transparent marks to help show the detail of the milestones while showing the overall picture.

```{r fig.width = 11, fig.height = 10}
ggplot(df_final, aes(x =  MilestoneDate, y = PrincipleName, col = MilestoneStatus, 
                fill = MilestoneStatus)) +
  geom_jitter(aes(alpha = 0.4, size = Weight),  width = 0.3, height = 0.3, shape = 21) +
  scale_x_date("Date", date_breaks = "3 months", date_labels = "%b %Y") +
  scale_y_discrete("Principle") +
  scale_fill_manual(values = c("green","royalblue2", "red"), labels = c("Active","Completed", "Overdue")) +
  scale_size(guide = FALSE) +
  scale_alpha(guide = FALSE) +
  geom_vline(xintercept = as.numeric(current_date), linetype = "dotted", col = "gray22") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  ggtitle(label = "Risk And Data Aggregation And Reporting - Project Progress",
          subtitle = "Shows milestones by principle, date and status. (More important milestones have a larger size)")
```
